---
title: "AB_test"
output: html_document
date: "2025-12-01"
---
read in data and library
```{r}
library(tidyverse)

clean_data<- read.csv("clean_data/HJ_data.csv")
```

AB test description:
- Null hypothesis: there is no difference in native bee abundance between between native and non-native plants.
- Alternative hypothesis: there is higher native bee abundance at native plants than non-native plants.

```{r}
##only testing native pollinator interactions right now so filter out non-native pollinators
#only keep columns relevant to this test- 
native_bees<- clean_data %>%
  filter(NativePollinator==1) %>%
  select(PollinatorGenusSpecies, PlantGenusSpecies, NativePlant)
#now all pollinators included are native so we don't need the NativePollinator column
```

Groups/treatments are native and non-native plants

- Test statistic is the difference in mean abundances of native bees observed at native vs. non-native plants

OR diff in abundances of native bees observed at native vs. non-native plants/abundance of those types of plants

Test statistic is the difference between the ratios of bee abundance at either native or non-native plants divided by the number of native or non-native species present

- Labels of native and non-native plants would be shuffled and test statistic recalculated to generate a histogram and p-value to determine if there is a significant difference between groups

##figure out how to calculate test stat- end with finding observed stat

test stat= number of bee interactions per species then take mean number of interactions per group
```{r}
## number of bee observations per plant species
# will use this dataframe for shuffling labels
abundance_by_plant<- native_bees %>%
  group_by(PlantGenusSpecies, NativePlant) %>%
  summarize(abundance= n(), .groups="drop")

## find average number of interactions per species based on group (native vs. non-native)
av_interactions<- abundance_by_plant %>%
  group_by(NativePlant) %>%
  summarize(mean_interactions= mean(abundance), .groups="drop")

#find difference between them
# average number of interactions for a native species - average number of interactions for a non-native species
observed_stat <- abs(av_interactions$mean_interactions[av_interactions$NativePlant=="1"] - av_interactions$mean_interactions[av_interactions$NativePlant=="0"])

observed_stat
```

- We will need a function to calculate the test statistic:
- - It will group by native and non-native plants to calculate the abundance of native bee visits at each.
- - It will then calculate the difference between those abundances and that is the test statistic
```{r}
####define function using above calculations
### this function calculates the observed statistic by first finding the abundance of native bees per plant species and then taking the average of those abundances for both native and non-native groups of plant species
calc_statistic<- function(df){
  #first finds the abundance of native bee interactions for each plant species
  df2<- df %>%
    group_by(PlantGenusSpecies, NativePlant) %>%
    summarize(abundance= n(), .groups="drop")
  #next finds the average abundance for species based on group (native or non-native plants)
  df3<- df2 %>%
    group_by(NativePlant) %>%
    summarize(mean=mean(abundance), .groups="drop")
  #finally calculates test statistic= the difference between average abundances of each group
  obs_stat <- abs(df3$mean[df3$NativePlant=="1"]- df3$mean[df3$NativePlant=="0"])
  return(obs_stat)
  }


## test it works- should match observed statistic above of 88.63158
calc_statistic(native_bees)
```

##figure out shuffling and running one test
```{r}
## for reproducibility
set.seed(123)

##shuffle label column from native_bees
native_bees_shuffle <-  sample(native_bees$NativePlant, 
                         nrow(native_bees), replace=FALSE)

## create data frame to use for test without altering og
native_bees_test<- native_bees

#create new column with shuffled labels
native_bees_test$NativePlantShuffled<- native_bees_shuffle

#check that the column was added
head(native_bees_test)

test2<- native_bees_test %>%
    group_by(PlantGenusSpecies, NativePlantShuffled) %>%
    summarize(abundance= n(), .groups="drop")

### new function for calculating test statistic with shuffled column instead of the original NativePlant column
### this function calculates the test statistic. It functions the same way as the calc_stat function except it uses the re-shuffled NativePlantShuffled labels instead of the original NativePlant labels that the first function used to calculate the observed statistic.
calc_stat_shuffled<- function(df){
  #first finds the abundance of native bee interactions for each plant species given a dataframe
  df2<- df %>%
    group_by(PlantGenusSpecies, NativePlantShuffled) %>%
    summarize(abundance= n(), .groups="drop")
  #next finds the average abundance for species based on group (native or non-native plants)
  df3<- df2 %>%
    group_by(NativePlantShuffled) %>%
    summarize(mean=mean(abundance), .groups="drop")
  obs_stat <- abs(df3$mean[df3$NativePlantShuffled=="1"]- df3$mean[df3$NativePlantShuffled=="0"])
  return(obs_stat)
  }

#test with native_bees after it's been shuffled once, run whole cell a few times to ensure this gives different numbers: it does
calc_stat_shuffled(native_bees_test)

```


- We will need a second function that uses the first function to run a permutation test:
- - It will reshuffle the native plant and non-native plant labels and re-run the above function to generate a new statistic
```{r}
## for reproducibility
set.seed(123)

#### function to run shuffling and test with calc_stat_shuffled
### this function creates a vector of shuffled labels from the NativePlant column from a given dataframe and adds that vector to a new column in the dataframe. It then calculates a test statistic given the newly re-shuffled lables in the dataframe using the calc-stat-shuffled function defined above.
run_one_test<- function(df){
  #first create a vector of shuffled labels (native or non-native plant) from the NativePlant column, without replacing values after they were sampled and making it the same length as the original NativePlant column so the same number of native and non-native labels are present
  shuffled_labels<- sample(df$NativePlant,
                           nrow(df), replace=FALSE)
  # next add the shuffled vector back to the dataframe
  df$NativePlantShuffled<- shuffled_labels
  # finally calculate the new test statistic from the shuffled labels using the calc_stat_shuffled function
  test_stat<- calc_stat_shuffled(df)
  return(test_stat)
}

## test function- run a couple times to ensure you get different results
run_one_test(native_bees)

```


- We will then need a for loop to run the AB test:
- - It will take the second function to reshuffle and re-run the test statistic over a number of loops to generate a vector of statistics
```{r}
## set up empty vector to fill with test statistic values
test_stats<- vector(mode="numeric", length=0)

#### create for loop
### this for loop runs 1000 permutation tests using the run_one_test function defined above and adds each value to an originally empty vector of test statistics
for (i in 1:1000){
  # first run a permuation test with the run_one_test function to get a test statistic back
  one_test<- run_one_test(native_bees)
  # next add that new test statistic to the test_stats vector
  test_stats<- append(test_stats, one_test)
}

## look at vector to see if it filled- it is
test_stats[1:10]

## test it contains numeric values- it does
class(test_stats)

```

find p-value
```{r}
## calculate p-value
# how probable that test stats are equal to observed or farther in direction of alt hypothesis
p_val <- mean(abs(test_stats)<= abs(observed_stat))
p_val
```

- We will then plot these statistics and compare the observed statistic to the histogram to see if there is a significant difference between the two groups (native and non-native plants).
```{r}
## turn test_stats vector into a dataframe to plot
test_stats_df<- as.data.frame(test_stats)

##plot histogram- results from for loop AND observed test statistic
ggplot(test_stats_df) +
  geom_histogram(aes(x=test_stats)) +
  geom_point(aes(x=observed_stat, y=0),color="red")
```


- This will also be for question 1.



